---
title: "My Awesome Title"
subtitle: "Daryn Ramsden"
author: "thisisdaryn at gmail dot com"
date: "last updated: `r Sys.Date()`"
institution: ""
output:
  xaringan::moon_reader:
    lib_dir: libs
    css: xaringan-themer.css
    chakra: libs/remark.js
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---



```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
knitr::opts_chunk$set(fig.align="center", fig.width=5, fig.height=5, warning = FALSE, message = FALSE)
```

```{r xaringan-themer, include = FALSE}
library(xaringanthemer)
duo_accent(
  primary_color = "ivory",
  secondary_color = "#310A31",
  header_font_google = google_font("Roboto", "400"),
  text_font_google   = google_font("Lato", "300"),
  code_font_family = "Fira Code",
  code_font_url = "https://cdn.rawgit.com/tonsky/FiraCode/1.204/distr/fira_code.css",
  header_color = "#f54278",
  title_slide_text_color = "#354a66"
)
```


### A note 

  * These slides make use of the <tt>chi_emps</tt> data set contained in the <tt>workshop<tt> package. 

  * 

```{r}
library(workshop)
data(chi_emps)
```



---
### Getting summary information

How big is the  data set? 
--
```{r}
dim(chi_emps)
```
--
What are the names of the columns?
```{r}
names(chi_emps)
```
--
You can use **summary** or **View** as well to get more info. 


---
### summary 

What happens when you run the following? 

```{r eval = FALSE}
summary(chi_emps)
```

---
### summary 

What happens when you run the following? 

```{r}
summary(chi_emps)
```


---
### (View)


and what about this?

```{r eval = FALSE}
View(chi_emps)
```



---
### Selecting only some columns

In base R, **[]** is used to subset data. For a 2d data structure, we have row conditions and column conditions separated by a comma inside the **[]**.
--


For example, we can select the first 5 rows and the first 4 columns of the *chi_emps* data frame:

--
```{r}
chi_emps[c(1:5),c(1:4)]
```



---
### Selecting columns by name

It's often more practical to select columns by name however. The following code:
  - keeps only the *Name*, *Dept*, *AnnSalary* columns
  - displays the dimensions of the smaller data set
--
```{r}
chi2 <- chi_emps[, c("Name", "Dept", 
                    "AnnSalary")]
dim(chi2)
```
--
Note that there was no row constraint in the first line as we were intending to keep all rows of the data.

---
### Rows can also be selected using logical operators

How do we select all the rows where the annual salary is in the interquartile range
--


The code below:
  - keeps only the rows where the salary is between than $75408 and $97440 
  - displays the first few rows
--
```{r}
midsal <- chi2[chi2$AnnSalary >= 75408 & chi2$AnnSalary <= 97440, ]
head(midsal)
```
--
Note that there is no column constraint as we are keeping all the columns.


---
### What are the other logical operators that can be used in filtering rows?

The previous slide made use of **<**, multiple other operators are available for filtering: 
--
```{r echo = FALSE}
Operator <- c( "==", "!=", "<", "<=",  ">", ">=", "|", "&", "!")
Meaning <- c("equal", "not equal", 
             "less than", "less than or equal to", 
              "greater than", "greater than or equal to",
             "Or: at least one of the expressions is true", 
             "And: Both expressions are true", 
             "Not: the expression is not true" )
optable <- data.frame(Operator, Meaning)
kableExtra::kable(optable)
```




---
### Some more examples

How would we get a data set with only hourly paid employees?

```{r}
hourly <- chi_emps[chi_emps$SalHour == "Hourly", ]
```




---
### Following up on that

How would we get loans that meet both of the following:
  - have an hourly rate more than $20/hr
  - typical hours more than 25 
--

```{r}
df <- hourly[(hourly$HourlyRate > 20) & 
               (hourly$TypicalHours > 25), ]
head(df)
```


---
### Another example

How would we get loans that meet either of the following:
  - the Department is *PUBLIC LIBRARY* 
  - the employee is Part-time, (<tt>FullPart</tt>: *P*) 
  
--
```{r}
lib_part <- chi_emps[chi_emps$Dept == "PUBLIC LIBRARY" | 
                chi_emps$FullPart == "P", ] 
head(lib_part)
```

  

---
### Using **$** to reference columns 

Another means of subsetting data that we would have seen before is the **$** 
--
```{r fig.height=5}
hist(chi_emps$HourlyRate)
```



--


We have been using it to extract single columns from the data


---
### **$** can be used to create new columns

We can also create new columns in the data set. 


For example, here's a new column that says whether or not the interest rate on a loan is greater than 10%:
--
```{r}
chi2$sal_gt100 <- ifelse(chi2$AnnSalary > 100000, TRUE, FALSE)
head(chi2)
```


---
### Explaining **ifelse**

The **ifelse** function is very handy and has three arguments:

- an expression to evaluate
- a result if the expression is TRUE
- a result if the expression is FALSE

