---
title: "dplyr 1: core functions"
subtitle: "Daryn Ramsden"
author: "thisisdaryn at gmail dot com"
date: "last updated: `r Sys.Date()`"
institution: ""
output:
  xaringan::moon_reader:
    lib_dir: libs
    css: xaringan-themer.css
    chakra: libs/remark.js
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---



```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
knitr::opts_chunk$set(fig.align="center", fig.width=5, fig.height=5, warning = FALSE, message = FALSE)
```

```{r xaringan-themer, include = FALSE}
library(xaringanthemer)
duo_accent(
  primary_color = "ivory",
  secondary_color = "#310A31",
  header_font_google = google_font("Roboto", "400"),
  text_font_google   = google_font("Lato", "300"),
  code_font_family = "Fira Code",
  code_font_url = "https://cdn.rawgit.com/tonsky/FiraCode/1.204/distr/fira_code.css",
  header_color = "#f54278",
  title_slide_text_color = "#354a66"
)
```

### What is dplyr?


  * subsetting
  
    * <tt>filter</tt>: selecting specified rows 
    * <tt>select</tt>: selecting specified columns
    
--

  * sorting
  
    * <tt>arrange</tt>
    
--


  * adding new variables
    
    * <tt>mutate</tt>
--

  * aggregating data
  
    * <tt>group_by</tt> and <tt>summarise</tt>



---
### The common pattern 

Each of these functions meet the following:
  
  * the first argument is a data frame
  
  * there are a variable number of following arguments
  
  
---
### filter: specifying row conditions

How do we get:
  * all individuals in either the Police or Fire Departments 

--

```{r}
library(workshop)
library(dplyr)
chi2 <- filter(chi_emps, Dept %in% c("POLICE", "FIRE"))
```

---
### filter

How do we get all individuals meeting the following:
  * in either the Police or Fire Departments
  * salaried

```{r}
chi3 <- filter(chi_emps, Dept %in% c("POLICE", "FIRE"),
               SalHour == "Salary")
```


---
### select

What if we wanted to keep only the columns of **chi3** related to salaried employees 

```{r}
chi4 <- select(chi3, Name, Titles,  Dept, FullPart, AnnSalary)
```


#### or 

--

```{r}
chi5 <- select(chi3, -SalHour, -HourlyRate, -TypicalHours)
```


--



Are <tt>chi4</tt> and <tt>chi5</tt> equivalent data sets? 



---
### arrange

How do we sort by salary?

```{r}
chi6 <- arrange(chi4, AnnSalary)
```

```{r}
chi7 <- arrange(chi4, desc(AnnSalary))
```


---
### group_by and summarise

  * <tt>group_by</tt>: partitions the data frame into groups by adding some invisible group information
  
  * <tt>summarise</tt>: aggregates data for each group 
  
  
---
### Example

What is the median salary in each department? 


```{r}
grp_sal <- group_by(chi_emps, Dept)
```



```{r}
avg_sals <- summarise(grp_sal, 
                AvgSalary = mean(AnnSalary, na.rm = TRUE))
```



```{r}
avg_sals <- summarise(grp_sal, 
                      AvgSalary = mean(AnnSalary, na.rm = TRUE),
                      numEmps = n())
```


---
### What 

```{r echo = FALSE}
reactable::reactable(avg_sals, )
```




